A Tale of Two Dictionary Learning Problems

### 2020-08-05 manuscripts/thesis/introduction.tex
new intro

It is a longstanding practice in the field of signal processing to describe signals as linear combinations of elementary waveforms from a pre-specified ``dictionary". When this dictionary forms a basis for the signal space, every signal has a unique decomposition into these atomic components. In the simplest such case, the basis is orthonormal and the representational coefficient scaling a given elementary waveform is merely the inner product of that waveform with the signal. %When non-orthogonal, the coefficients are computed by taking the inner product with respect to the components of the dictionary inverse.

@@

The need for greater freedom of expression eventually led to the development of redundant signal representations utilizing overcomplete dictionaries containing more atoms than there are dimensions of the signal. In this case, there are infinitely many ways in which a signal may be decomposed into its constituent components, and the intention is to seek the most informative such representation as measured by some task-specific cost function.

@@

Early approaches to sparse representation modeling assumed a model of the signal class from which a suitable sparsifying dictionary could then be derived, as we have in our recurring example. While such dictionaries are typically characterized by an analytic formulation and a fast implicit implementation, they unfortunately tend to be over-simplistic models when applied to natural phenomena.

An alternative modern approach to dictionary design is conditioned on the assumption that the sparse structure of signals conveying information about complex natural phenomena can be more accurately extracted directly from a training dataset, a process referred to as dictionary learning  (see \cite{Zhang15} for a comprehensive review). In the seminal work \cite{Olshausen96}, a dictionary trained over a collection of small patches extracted from images of the natural environment was shown to share qualitative similarities with linear filters estimated from the response properties of simple-cell neurons in mammalian visual cortex, which until then had been more weakly described analytically as Gabor filters (see also \cite{hurri1996image, bell1997independent, van1998independent}). This remarkable discovery showed that the assumption of sparsity alone could potentially account for a fundamental property of the visual system, and demonstrated the potential of the machine learning approach to dictionary design. Even more curiously, these waveforms (e.g., Gabor-like wavelets) have since tended to appear in dictionaries learned by a variety of machine learning algorithms trained over different natural image datasets, suggesting that the optimal dictionaries for these signals may, in some sense, be canonical \cite{donoho2001can}.

@@

%and more recent work relating compressed sensing \cite{eldar2012compressed} to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, 
It is natural to wonder, in light of these discoveries, when a dictionary admitting optimally sparse representations of a signal class is indeed identifiable from a representative sample. Answers to this question have implications in practice whenever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}). 
Even though several algorithms have recently been proposed to provably recover a unique dictionary under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state of the art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. Surprisingly, despite the now ubiquitous application of dictionary learning methods in practice, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data (noise being inevitable in practice) has yet, to the best of my knowledge, to appear in the literature.

@@

Let us proceed to rigorously define the two formulations of the dictionary learning problem with which this thesis is concerned. Fix a matrix $\mathbf{A} \in \mathbb R^{n \times m}$ with the elementary waveforms of the dictionary as its columns  $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k<m$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our worst-case uncertainty in measuring the product $\mathbf{A}\mathbf{x}_i$. We shall first consider the following decidable\footnote{Note that Prob.~\ref{InverseProblem} is decidable for rational inputs $\mathbf{z}_i$ \cite{chrishillar} since the statement that it has a solution can be expressed as a logical sentence in the theory of algebraically closed fields, and this theory has quantifier elimination \cite{basu2006algorithms}.} formulation of the dictionary learning problem.
%The first mathematical problem we consider is the following.

@@

Formal statements of the main findings described above are given in Chap.~\ref{Results}, along with their adaptation to dictionaries and codes drawn from arbitrary (continuous) probability distributions (Cor.~\ref{ProbabilisticCor}). For clarity of exposition, the technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt} are deferred to Chap.~\ref{DUT}, following some necessary definitions and the statement of a key lemma in combinatorial matrix analysis (Lem.~\ref{MainLemma}; proven in the Appendix). These results and their applications are discussed in Chap.~\ref{Discussion}, which concludes with some open questions and directions for future research seeded in part by the results of more practically-minded simulations. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

>>>>

It is a longstanding practice in the field of signal processing to describe signals as linear combinations of elementary waveforms selected from a pre-specified ``dictionary". When this dictionary forms a basis for the signal space, every signal has a unique decomposition into these atomic components. In the most basic such case, the basis is orthonormal and the representational coefficient scaling a given elementary waveform is simply the inner product of that waveform with the signal. %When non-orthogonal, the coefficients are computed by taking the inner product with respect to the components of the dictionary inverse.

@@

The need for such freedom of expression eventually led to the development of redundant signal representations utilizing overcomplete dictionaries, which contain more atoms than there are dimensions of the signal. In such cases, there are infinitely many ways in which a signal may be decomposed into a superposition of the waveforms constituting the dictionary, and the intention is rather to seek the most informative such representation as measured by some task-specific cost function.

@@

Early approaches to sparse representation modeling assumed a model of the signal class from which a suitable sparsifying dictionary could be derived, as we have in our recurring example. While such dictionaries are typically characterized by an analytic formulation and a fast implicit implementation, they unfortunately tend to be over-simplistic models when applied to natural phenomena.

An alternative modern approach to dictionary design is conditioned on the assumption that the sparse structure of signals conveying information about complex natural phenomena can be more accurately extracted directly from a training dataset, a process referred to as dictionary learning  (see \cite{Zhang15} for a comprehensive review). In the seminal work \cite{Olshausen96}, a dictionary trained over a collection of small patches extracted from images of the natural environment was shown to share qualitative similarities with linear filters estimated from the response properties of simple-cell neurons in mammalian visual cortex, which until then had been more weakly described analytically as Gabor filters (see also \cite{hurri1996image, bell1997independent, van1998independent}). This remarkable discovery showed that the assumption of sparsity alone could potentially account for a fundamental property of the visual system, and demonstrated the potential of the machine learning approach to dictionary design. Even more curiously, these waveforms (e.g., Gabor-like wavelets) tend to appear in dictionaries learned by a variety of machine learning algorithms trained over different natural image datasets, suggesting that the optimal dictionaries for these signals may, in some sense, be canonical \cite{donoho2001can}.

@@

%and more recent work relating compressed sensing \cite{eldar2012compressed} to a theory of information transmission through random wiring bottlenecks in the brain \cite{Isely10}, 
In light of these discoveries, it is natural to wonder when a dictionary admitting optimally sparse representations of a signal class is indeed identifiable from a representative sample. Answers to this question have implications in practice whenever an appeal is made to latent sparse structure of data (e.g., forgery detection \cite{hughes2010, olshausen2010applied}; brain recordings \cite{jung2001imaging, agarwal2014spatially, lee2016sparse}; and gene expression \cite{wu2016stability}). 
Even though several algorithms have recently been proposed to provably recover a unique dictionary under specific conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state of the art), few theorems can be invoked to justify the consistency of inference under this model of data more broadly. Surprisingly, despite the now ubiquitous application of dictionary learning methods in practice, a universal guarantee of the uniqueness and stability of learned dictionaries and the sparse representations they induce over noisy data (noise being inevitable in practice) has yet, to the best of my knowledge, to appear in the literature.

@@

Let us proceed to rigorously define the two formulations of the dictionary learning problem with which this work concerns itself. Fix a matrix $\mathbf{A} \in \mathbb R^{n \times m}$ with the elementary waveforms of the dictionary as its columns  $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k<m$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our worst-case uncertainty in measuring the product $\mathbf{A}\mathbf{x}_i$. We shall first consider the following decidable\footnote{Note that Prob.~\ref{InverseProblem} is decidable for rational inputs $\mathbf{z}_i$ \cite{chrishillar} since the statement that it has a solution can be expressed as a logical sentence in the theory of algebraically closed fields, and this theory has quantifier elimination \cite{basu2006algorithms}.} formulation of the dictionary learning problem.
%The first mathematical problem we consider is the following.

@@

Formal statements of the main findings described above are given in Chap.~\ref{Results}, along with their adaptation to dictionaries and codes drawn from arbitrary (continuous) probability distributions (Cor.~\ref{ProbabilisticCor}). All results assume real matrices and vectors. For clarity of exposition, the technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt} are deferred to Chap.~\ref{DUT}, following some necessary definitions and the statement of a key lemma in combinatorial matrix analysis (Lem.~\ref{MainLemma}, the proof of which is relegated to the chapter's Appendix). These results and their applications are discussed in Chap.~\ref{Discussion}, which concludes with some open questions and directions for future research seeded in part by the results of more practically-minded simulations. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

### 2020-08-05 manuscripts/thesis/proofs.tex
new intro

% WITHOUT J SUBSCRIPT

@@

The required stability result in combinatorial matrix analysis is the following. For expository purposes, the proof of this fact is relegated to the Appendix.

@@

The constant $C_2$ (a function of $\mathbf{A}$ and $\mathcal{H}$) will be presented relative to a quantity used in \cite{Deutsch12} to analyze the convergence of the ``alternating projections" algorithm for projecting a point onto the intersection of subspaces. This quantity is incorporated into the following definition, which we shall refer to in the proof of Lem.~\ref{DistanceToIntersectionLemma} in the Appendix; specifically, it will be used to bound the distance between a point and the intersection of subspaces given an upper bound on its distance from each subspace.

@@

Several results in the literature had to be combined to extend the guarantees derived in \cite{Hillar15} into the noisy regime. The main challenge was to generalize Lem.~\ref{MainLemma} to the case where the $k$-dimensional subspaces spanned by corresponding submatrices $\mathbf{A}_S$ and $\mathbf{B}_{\pi(S)}$ are assumed to be``close" but not identical. Referring now to the proof in the Appendix, the situation is unlike that in \cite{Hillar15}, where an inductive argument could be applied to the noiseless case. Rather, here it has to be explicitly demonstrated that this proximity relation is propagated through iterated intersections right down to the spans of the dictionary elements themselves. Lem.~\ref{DistanceToIntersectionLemma} was designed to encapsulate this fact, proven by appeal to a convergence guarantee for an alternating projections algorithm first proposed by von Neumann. This result, combined with a little known fact \eqref{eqdim} about the distance metric between subspaces, make up the more obscure components of the deduction.

>>>>

% WITHOUT J SUBSCRIPT
We shall prove Thm.~\ref{DeterministicUniquenessTheorem} and then prove Thm.~\ref{SLCopt} by arguing that Prob.~\ref{OptimizationProblem} reduces to an instance of Prob.~\ref{InverseProblem} given sufficient data.

@@

The required stability result in combinatorial matrix analysis is the following. For expository purposes, the proof of this fact is relegated to Sec.~\ref{proofs}.

@@

The constant $C_2$ (a function of $\mathbf{A}$ and $\mathcal{H}$) will be presented relative to a quantity used in \cite{Deutsch12} to analyze the convergence of the ``alternating projections" algorithm for projecting a point onto the intersection of subspaces. This quantity is incorporated into the following definition, which we shall refer to in the proof of Lem.~\ref{DistanceToIntersectionLemma} in the Sec.~\ref{proofs}; specifically, it will be used to bound the distance between a point and the intersection of subspaces given an upper bound on its distance from each subspace.

@@

Several results in the literature had to be combined to extend the guarantees derived in \cite{Hillar15} into the noisy regime. The main challenge was to generalize Lem.~\ref{MainLemma} to the case where the $k$-dimensional subspaces spanned by corresponding submatrices $\mathbf{A}_S$ and $\mathbf{B}_{\pi(S)}$ are assumed to be``close" but not identical. Referring now to the proof in this chapter's Appendix, the situation is unlike that in \cite{Hillar15}, where an inductive argument could be applied to the noiseless case. Rather, here it has to be explicitly demonstrated that this proximity relation is propagated through iterated intersections right down to the spans of the dictionary elements themselves. Lem.~\ref{DistanceToIntersectionLemma} was designed to encapsulate this fact, proven by appeal to a convergence guarantee for an alternating projections algorithm first proposed by von Neumann. This result, combined with a little known fact \eqref{eqdim} about the distance metric between subspaces, make up the more obscure components of the deduction.

### 2020-08-05 manuscripts/thesis/results.tex
new intro

A precise statement of the results requires first the identification of some combinatorial criteria on the supports\footnote{Recall that a vector $\mathbf{x}$ is said to be \emph{supported} in $S$ when $\mathbf{x} \in \text{\rmfamily span}\{\mathbf{e}_j: j\in S\}$, with $\mathbf{e}_j$ forming the standard column basis.} of sparse vectors. Let $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$.  A \emph{hypergraph} on vertices $[m]$  is simply any subset $\mathcal{H} \subseteq 2^{[m]}$. Let us say that $\mathcal{H}$ is \textit{$k$-uniform} when $\mathcal{H} \subseteq {[m] \choose k}$. The \emph{degree} $\deg_\mathcal{H}(i)$ of a node $i \in [m]$ is the number of sets in $\mathcal{H}$ that contain $i$, and we say $\mathcal{H}$ is \emph{regular} when for some $r$ we have $\deg_\mathcal{H}(i) = r$ for all $i$ (given such an $r$, we say $\mathcal{H}$ is \textit{$r$-regular}). Let us also write $2\mathcal{H} := \{ S \cup S': S, S' \in \mathcal{H}\}$.  The following class of structured hypergraphs is a key ingredient in this work.

@@

Next, a quantitative generalization of the spark condition (\ref{SparkCondition}) to combinatorial subsets of supports is given. The \emph{lower bound} of an $n \times m$ matrix $\mathbf{M}$ is the largest $\alpha$ with \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$ \cite{Grcar10}. By compactness of the unit sphere, every injective linear map has a positive lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then submatrices formed from $2k$ of its columns or less have strictly positive lower bounds.

@@

We are now in a position to state the main result, though for expository purposes the quantity $C_1$ will be left%(a function of $\mathbf{A}$, $\mathbf{x}_1, \ldots, \mathbf{x}_N$, and $\mathcal{H}$) 
undefined until Chap.~\ref{DUT}.
 All results  below assume real matrices and vectors.

@@

It is befitting to conclude this chapter with some comments on the optimality of these results.  The linear scaling for $\varepsilon$ in \eqref{epsdel} is essentially optimal (e.g., see \cite{arias2013fundamental}), but a basic open problem remains: how many samples are necessary to determine the sparse coding model? 
These results demonstrate that sparse codes $\mathbf{x}_i$ drawn from only a polynomial number of $k$-dimensional subspaces permit stable identification of the generating dictionary $\mathbf{A}$. 
This lends some legitimacy to the use of the model in practice, where data in general are unlikely (if ever) to exhibit the exponentially many possible $k$-wise combinations of dictionary elements required by (to my knowledge) all previously published results.

@@

As it seemed to benefit a reviewer of this work, some clarification may be in order on how the deterministic sample complexity $N = |\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ given here compares to those listed in the the top two rows listed in Table I of \cite{Hillar15}. To be clear, the theory developed here (Cor.~\ref{DeterministicUniquenessCorollary}) is strictly more general, since $\mathcal{H}$ can always be taken to be ${[m] \choose k}$. The point of difference in this comparison is the assumed set of supports for sparse codes, which is always ${[m] \choose k}$ in \cite{Hillar15}, whereas here it can be assumed to be any regular $k$-uniform hypergraph that satisfies the SIP. By row, %, assuming for our results that sufficient data have been sampled from all supports in a hypergraph H satisfying the SIP:
\begin{enumerate}
\item[I.] The result here improves upon the listed $k{m \choose k}^2$ by an exponential factor, since for every $k < m$ there exists a regular $k$-uniform hypergraph $\mathcal{H}$ with $|\mathcal{H}| = m$ satisfying the SIP. 
\item[II.] The authors in \cite{Hillar15} have applied measure-theoretic arguments to achieve $(k+1){m \choose k}$ with almost-certainty (i.e. with probability one), a factor of $m$ reduction over that for which certainty can alternatively be guaranteed here.    
%\item[III.] A direct comparison cannot be made because it is an open question as to the probability that a random set of supports satisfies the SIP. This would be an interesting problem for the community to solve; regardless, the result of \cite{Hillar15} is still implied by the more general theory given here.
%\item[IV.] The result here is $|\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ with probability one compared to $(k+1){m \choose k}$ with probability one. In this case the sample complexity stated in \cite{Hillar15} is marginally better, but it relies on a noise-free technicality unique to their assumption $\mathcal{H} = {m \choose k}$. %differs in flavor: their $\mathbf{x_i}$ are assumed to be distributed as $(k+1)$ samples per support for every support of size $k \in [m]$, whereas here it is assumed that $(k-1){m \choose k}$ codes $\mathbf{x_i}$ are distributed over each support in some hypergraph $\mathcal{H}$ satisfying the SIP. 
%\item[V.] Again, a direct comparison cannot be made for the reason stated for row III.
\end{enumerate}

>>>>

To state precisely the results of this work, we will require first the identification of some combinatorial criteria on the supports\footnote{Recall that a vector $\mathbf{x}$ is said to be \emph{supported} in $S$ when $\mathbf{x} \in \text{\rmfamily span}\{\mathbf{e}_j: j\in S\}$, with $\mathbf{e}_j$ forming the standard column basis.} of sparse vectors. Let $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$.  A \emph{hypergraph} on vertices $[m]$  is simply any subset $\mathcal{H} \subseteq 2^{[m]}$. Let us say that $\mathcal{H}$ is \textit{$k$-uniform} when $\mathcal{H} \subseteq {[m] \choose k}$. The \emph{degree} $\deg_\mathcal{H}(i)$ of a node $i \in [m]$ is the number of sets in $\mathcal{H}$ that contain $i$, and we say $\mathcal{H}$ is \emph{regular} when for some $r$ we have $\deg_\mathcal{H}(i) = r$ for all $i$ (given such an $r$, we say $\mathcal{H}$ is \textit{$r$-regular}). Let us also write $2\mathcal{H} := \{ S \cup S': S, S' \in \mathcal{H}\}$.  The following class of structured hypergraphs is a key ingredient in this work.

@@

Next, we will require a quantitative generalization of the spark condition (\ref{SparkCondition}) to combinatorial subsets of supports. The \emph{lower bound} of an $n \times m$ matrix $\mathbf{M}$ is the largest $\alpha$ with \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$ \cite{Grcar10}. By compactness of the unit sphere, every injective linear map has a positive lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then submatrices formed from $2k$ of its columns or less have strictly positive lower bounds.

@@

The following is the statement of the main result. For expository purposes the quantity $C_1$ (a function of $\mathbf{A}$, $\mathbf{x}_1, \ldots, \mathbf{x}_N$, and $\mathcal{H}$) will be left undefined until Chap.~\ref{DUT}.

@@

It is befitting to comment on the optimality of these results. The linear scaling for $\varepsilon$ in \eqref{epsdel} is essentially optimal (e.g., see \cite{arias2013fundamental}), but a basic open problem remains: how many samples are necessary to determine the sparse coding model? 
These results demonstrate that sparse codes $\mathbf{x}_i$ drawn from only a polynomial number of $k$-dimensional subspaces permit stable identification of the generating dictionary $\mathbf{A}$. 
This lends some legitimacy to the use of the model in practice, where data in general are unlikely (if ever) to exhibit the exponentially many possible $k$-wise combinations of dictionary elements required by (to my knowledge) all previously published results.

@@

As it seemed to benefit a reviewer of this work, some clarification may be in order on how the deterministic sample complexity $N = |\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ given here (Cor.~\ref{DeterministicUniquenessCorollary}) compares to those listed in the the top two rows listed in Table I of \cite{Hillar15}. To be clear, the theory developed here is strictly more general, since $\mathcal{H}$ can always be taken to be ${[m] \choose k}$. The point of difference in this comparison is the assumed set of supports for sparse codes, which is always ${[m] \choose k}$ in \cite{Hillar15}, whereas here it can be assumed to be any regular $k$-uniform hypergraph that satisfies the SIP. By row, %, assuming for our results that sufficient data have been sampled from all supports in a hypergraph H satisfying the SIP:
\begin{enumerate}
\item[I.] The result here improves upon the listed $k{m \choose k}^2$ by an exponential factor, since for every $k < m$ there exists a regular $k$-uniform hypergraph $\mathcal{H}$ with $|\mathcal{H}| = m$ satisfying the SIP. 
\item[II.] The authors in \cite{Hillar15} have applied measure-theoretic arguments to achieve $(k+1){m \choose k}$ with almost-certainty (i.e. with probability one), a factor of $m$ reduction over that for which certainty can alternatively be guaranteed here.    
%\item[III.] A direct comparison cannot be made because it is an open question as to the probability that a random set of supports satisfies the SIP. This would be an interesting problem for the community to solve; regardless, the result of \cite{Hillar15} is still implied by the more general theory given here.
%\item[IV.] The result here is $|\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ with probability one compared to $(k+1){m \choose k}$ with probability one. In this case the sample complexity stated in \cite{Hillar15} is marginally better, but it relies on a noise-free technicality unique to their assumption $\mathcal{H} = {m \choose k}$. %differs in flavor: their $\mathbf{x_i}$ are assumed to be distributed as $(k+1)$ samples per support for every support of size $k \in [m]$, whereas here it is assumed that $(k-1){m \choose k}$ codes $\mathbf{x_i}$ are distributed over each support in some hypergraph $\mathcal{H}$ satisfying the SIP. 
%\item[V.] Again, a direct comparison cannot be made for the reason stated for row III.
\end{enumerate}
